---
title: "Health and Wealth Around the World"
subtitle: "Gapminder data, 1952–2007"
date: "2025-10-16"
bibliography: bib/references.bib
csl: bib/chicago-author-date.csl
format:
  html:
    code-fold: false
    code-summary: "Show code"
    page-layout: full
    toc: false
    knitr:
      opts_chunk: 
        fig-align: "center"
execute:
  warning: false
  message: false
  echo: false
---

```{r}
#| label: load-libraries-data
#| warning: false
#| message: false

library(tidyverse)
library(patchwork)
library(ggridges)
library(tinytable)

theme_gapminder <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "bold", color = "gray50"),
    axis.title.x = element_text(hjust = 0),
    axis.title.y = element_text(hjust = 1)
  )

gapminder <- read_csv("data/gapminder.csv")

n_countries <- n_distinct(gapminder$country)
first_year <- min(gapminder$year)
last_year <- max(gapminder$year)
```

## Introduction

This analysis shows trends in life expectancy and GDP per capita for `{r} n_countries` countries from XXXX to YYYY. The data was originally collected by Hans Rosling and the Gapminder foundation.

## Continent-level trends

Average life expectancy increased substantially between XXXX and YYYY. Asia saw the biggest average increase (see @tbl-continent).

```{r}
#| label: fig-distribution-continent
#| fig-cap: "Distribution of life expectancy within continents"
#| fig-width: 6
#| fig-height: 4
#| out-width: 80%

lifeexp_1952 <- gapminder |>
  filter(year == 1952, continent != "Oceania") |>
  ggplot(aes(x = lifeExp, y = fct_rev(continent), fill = continent)) +
  geom_density_ridges(color = "white") +
  scale_x_continuous(limits = c(20, 90)) +
  guides(fill = "none") +
  labs(
    x = NULL,
    y = NULL,
    title = "Life expectancy by continent",
    subtitle = "1952"
  ) +
  theme_gapminder

lifeexp_2007 <- gapminder |>
  filter(year == 2007, continent != "Oceania") |>
  ggplot(aes(x = lifeExp, y = fct_rev(continent), fill = continent)) +
  geom_density_ridges(color = "white") +
  scale_x_continuous(limits = c(20, 90)) +
  guides(fill = "none") +
  labs(
    x = "Life expectancy (years)",
    y = NULL,
    subtitle = "2007",
    caption = "Oceania omitted"
  ) +
  theme_gapminder

lifeexp_1952 / lifeexp_2007
```

```{r}
#| label: tbl-continent
#| tbl-cap: "Change in average life expectancy by continent"
#| message: false

gapminder |>
  filter(year %in% c(1952, 2007)) |>
  group_by(continent, year) |>
  summarize(avg_life = mean(lifeExp)) |>
  group_by(continent) |>
  mutate(diff = avg_life - lag(avg_life)) |>
  ungroup() |>
  pivot_wider(names_from = year, values_from = c(avg_life, diff)) |>
  arrange(desc(diff_2007)) |>
  select(
    Continent = continent,
    `1952` = avg_life_1952,
    `2007` = avg_life_2007,
    Change = diff_2007
  ) |> 
  tt(digits = 3) |> 
  style_tt(j = 1, align = "l") |> 
  style_tt(j = 2:4, align = "c") |> 
  style_tt(i = 1, color = "#03BFC4", bold = TRUE) |> 
  group_tt(j = list(Averages = 2:3))
```

## Country-level trends

@fig-health-wealth recreates a part of Hans Rosling's famous ["200 countries, 200 years, 4 minutes" video](https://www.youtube.com/watch?v=jbkSRLYSojo) and shows the relationship betweeen GDP per capita and life expectancy.

```{r}
#| label: fig-health-wealth
#| fig-cap: "GDP per capita and life expectancy are closely related"
#| fig-width: 6
#| fig-height: 4
#| out-width: 80%

gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_smooth(method = "lm", color = "orange") +
  geom_point(aes(size = pop, color = continent)) +
  scale_x_log10(labels = scales::label_dollar(accuracy = 1)) +
  scale_size_continuous(labels = scales::label_comma()) +
  labs(
    x = "GDP per capita",
    y = "Life expectancy (years)",
    size = "Country population",
    color = "Continent",
    title = "Health and wealth in 2007"
  ) +
  theme_gapminder
```
